Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-08 paper/current/main.tex
formatting author block
adjust supplemental formatting
edit intro, update results

% CUSTOM MACROS

@@

\begin{document} %{{{
\title{Unsupervised embedding of trajectories captures the latent structure of mobility} %{{{
\author[1,4]{Dakota Murray}
\author[1, 2, 4]{Jisung Yoon}
\author[1]{Sadamori Kojaku}
\author[3]{Rodrigo Costas}
\author[2]{Woo-Sung Jung}
\author[1]{Staša Milojević \thanks{smilojev@indiana.edu}}
\author[1]{Yong-Yeol Ahn \thanks{yyahn@iu.edu}}

\affil[1]{Center for Complex Networks and Systems Research, Luddy School of Informatics, Computing, and Engineering, Indiana University, Bloomington, IN 47408, USA}
\affil[2]{Department of Industrial and Management Engineering, Pohang University of Science and Technology,
Pohang 37673, Republic of Korea.}
\affil[3]{Centre for Science and Technology Studies (CWTS), Leiden University, P.O. Box 905,
2300 AX Leiden, The Netherlands}
\affil[4]{These authors contributed equally: Jisung Yoon, Dakota Murray.}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Historically, human mobility is strongly constrained by physical separation---geographical distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
The critical importance of understanding mobility in many socio-economic phenomena thus compel alternative frameworks for mobility.
Here, using a massive dataset of scientific mobility, U.S. flight itineraries and Korean accommodation reservation, we demonstrate that a neural embedding model~\autocite{mikolov2013word2vec} can encode nuanced relationships between locations into a vector-space, providing an effective functional distance that reflects the multi-faced structure of human mobility. 
Focusing on the case of scientific mobility, we show that embeddings of scientific organizations discovers cultural and linguistic relations at multiple levels of granularity, and even academic prestige.
Furthermore, the embedding vectors themselves reveal universal relationships between organizational characteristics, and their place in the global landscape of scientific mobility. 
The ability to learn scalable, dense and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which we can not only derive a meaningful \textit{functional distance} between places but also probe relationships based on culture, language, and even prestige along with geographic relationship. 
We leverage three massive datasets: a dataset of U.S. passenger flight itinerary records, Korean accommodation reservation records, and a dataset of scientists' career mobility between organizations captured in bibliometric records. 
In particular, we focus on scientific mobility due to its richness and importance. Scientific mobility, which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic in the Science of Science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

The flight itinerary data, from the Airline Origin and Destination Survey, consists of records of more than 300 million itineraries between 1993 and 2020 documenting domestic flights between 828 airports in the United States. 
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports. 
The Korean accommodation reservation records consist of accommodation reservation history between 2018 and 2020 between 1,038 unique accommodation in Seoul, South Korea.
A trajectory is constructed for each customer's accommodation reservation history, constructing an ordered sequence of unique commendation with accurate location information.
Although more complex, trajectories of scientific mobility are constructed in a similar fashion. 
Trajectories are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
Each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).

@@

We show that the embedding distance, compared with the geographical distance, predicts actual mobility flows. 
In the case flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b). 
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Methods) and computed the expected flux  $\hat{T}_{ij}$ using the geographic distance and the embedding distance as parameters.
The embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}c-d).
These results are consistent in the case of scientific mobility. 
The amount of variance in the expected flux explained by the embedding distance ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) is again more than double that explained by the geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}f).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
As with flight itineraries, we fit the gravity model on the scientific mobility data using the gravity model.
Predictions made using the embedding distance outperform those made using the geographic distance (Fig.~\ref{fig:gravity_performance}g-h), which also holds for the subsets of only domestic and only international mobility (Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:predict_emb_bymodel}).
For scientific mobility, the embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results demonstrate that the embedding distance better captures patterns of actual mobility than does the geographic distance across datasets.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts U.S flights and global scientific mobility.}
		\textbf{a-b.} Embedding distance (cosine distance between organization vectors) better explains the expected flux of passengers between U.S. airports than does geographic distance. 
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{c-d.} Predictions of flux between airport pairs made using embedding distance outperform those made using geographic distance. 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility. 
		\textbf{e-f.} Embedding distance better explains the expected flux of researchers between organization pairs than geographic distance.
		\textbf{g-h.} Predictions of flux between organization pairs made using embedding distance outperform those made using geographic distance.
	}
	\label{fig:gravity_performance}
\end{figure}

@@

\subsection*{U.S. flight itinerary data}
We source U.S. airport itinerary data from the Origin and Destination Survey (DB1B), provided by the Bureau of Transportation Statistics at the United States Department of Transportation. 
DB1B is a sample of 10 percent of domestic airline tickets between 1993 and 2020, comprising 307,760,841 passenger itineraries between 828 U.S. airports. 
Each itinerary is associated with an origin, destination, and passenger and ticket metadata.

\subsection*{Korean accommodation reservation data}
We source Korean accommodation reservation data from collaboration with Good Choice Company in Korea. Each trajectory is the sequence of reserved accommodation in Seoul, South Korea between 2018.08.01 ~ 2020.07.31, comprising 938,962 trajectories consist of  1,038 unique accommodations.

@@

@@

% S1 Table
\begin{table}
\caption{\textbf{Full organization names}}
\label{table:supp:orglabels}
\end{table}

% S2 Table
\begin{table}
\caption{\textbf{L2 Norm of country's representative vectors}.}
\label{table:supp:norm}
\end{table}

%
% Figure - publications over time
%
\begin{figure}[]
\begin{flushleft}
	%\includegraphics[width=\textwidth]{\figdir/Descriptive/pubs_over_time.pdf}
	\caption{
		\textbf{Publications over time.}
		}
\end{flushleft}
	\label{fig:supp:pubs_over_time}
\end{figure}

>>>>

% CUSTOM MACROS
\usepackage{ marvosym }

@@

\renewcommand\Authfont{\fontsize{12}{14.4}\selectfont}
\renewcommand\Affilfont{\fontsize{9}{10.8}\itshape}

\begin{document} %{{{
\title{Unsupervised embedding of trajectories captures the latent structure of mobility} %{{{
\author[1,\Yinyang]{Dakota Murray}
\author[1,2,\Yinyang]{Jisung Yoon}
\author[1]{Sadamori Kojaku}
\author[3]{Rodrigo Costas}
\author[2]{Woo-Sung Jung}
\author[1]{Staša Milojević \thanks{smilojev@indiana.edu}}
\author[1]{Yong-Yeol Ahn \thanks{yyahn@iu.edu}}

\affil[1]{Center for Complex Networks and Systems Research, Luddy School of Informatics, Computing, and Engineering, Indiana University, Bloomington, IN 47408, USA}
\affil[2]{Department of Industrial and Management Engineering, Pohang University of Science and Technology,
Pohang 37673, Republic of Korea.}
\affil[3]{Centre for Science and Technology Studies (CWTS), Leiden University, P.O. Box 905,
2300 AX Leiden, The Netherlands}
\affil[\Yinyang]{These authors contributed equally: Jisung Yoon, Dakota Murray.}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Historically, human mobility is strongly constrained by physical separation---geographical distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
The critical importance of understanding mobility in many socio-economic phenomena thus compel alternative frameworks for mobility.
Here, using a massive dataset of scientific mobility, U.S. flight itineraries, and South Korean accommodation reservations, we demonstrate that a neural embedding model~\autocite{mikolov2013word2vec} can encode nuanced relationships between locations into a vector-space, providing an effective functional distance that reflects the multi-faced structure of human mobility. 
Focusing on the case of scientific mobility, we show that embeddings of scientific organizations discovers cultural and linguistic relations at multiple levels of granularity, and even academic prestige.
Furthermore, the embedding vectors themselves reveal universal relationships between organizational characteristics, and their place in the global landscape of scientific mobility. 
The ability to learn scalable, dense and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which we can not only derive a meaningful \textit{functional distance} between places but also probe relationships based on culture, language, and even prestige along with geographic relationship. 
We leverage three massive datasets: U.S. passenger flight itinerary records, South Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records. 
In particular, we focus on scientific mobility due to its richness and importance. Scientific mobility, which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic in the Science of Science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

The flight itinerary data, from the Airline Origin and Destination Survey, consists of records of more than 300 million itineraries between 1993 and 2020 documenting domestic flights between 828 airports in the United States. 
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports. 
The Korean accommodation reservations consist of accommodation reservation history between 2018 and 2020 between 1,038 unique accommodation locations in Seoul, South Korea.
A trajectory is constructed for each customer's accommodation reservation history, constructing an ordered sequence of unique commendation with accurate location information.
Although more complex, trajectories of scientific mobility are constructed in a similar fashion. 
Trajectories are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
Each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).

@@

We show that the embedding distance, compared with the geographical distance, predicts actual mobility flows. 
In the case flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$). 
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Methods) and computed the expected flux  $\hat{T}_{ij}$ using the geographic distance and the embedding distance as parameters.
The embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
The neural embedding also performs well in a smaller geographic context:
In the case of South Korean hotel accommodations, such that the embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d). 
These results are consistent in the case of scientific mobility. 
The amount of variance in the expected flux explained by the embedding distance ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) is again more than double that explained by the geographic distance ($R^{2} = 0.22$).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
As with flight itineraries, we fit the gravity model on the scientific mobility data using the gravity model.
Predictions made using the embedding distance outperform those made using the geographic distance (Fig.~\ref{fig:gravity_performance}f), which also holds for the subsets of only domestic and only international mobility (Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:predict_emb_bymodel}).
For scientific mobility, the embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results demonstrate that the embedding distance better captures patterns of actual mobility than does the geographic distance across domains.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts U.S flights, accommodation reservations, and global scientific mobility.}
		\textbf{a} Embedding distance (cosine distance between organization vectors, top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom). 
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{b} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom). 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility. 
		For South Korean hotel reservations, embedding distance better explains the expected flux than does geographic distance (\textbf{c}), and produces better predictions (\textbf{d}).
		Similarly, in the case of global scientific mobility, embedding distance explains the expected flux between organizations (\textbf{e)} and allows for better predictions (\textbf{f}) than geographic distance.
	}
	\label{fig:gravity_performance}
\end{figure}

@@

\subsection*{U.S. flight itinerary data}
We source U.S. airport itinerary data from the Origin and Destination Survey (DB1B), provided by the Bureau of Transportation Statistics at the United States Department of Transportation. 
DB1B is a sample of 10 percent of domestic airline tickets between 1993 and 2020, comprising 307,760,841 passenger itineraries between 828 U.S. airports. 
Each itinerary is associated with a a trajectory of airports including an origin, destination, and intermediary stops.

\subsection*{Korean accommodation reservation data}
We source Korean accommodation reservation data from collaboration with Good Choice Company in Korea, containing 938,962 trajectories spanning the time period of August 2018 through July 2020,  comprising 1,038 unique accommodation locations in Seoul, South Korea.

@@

\clearpage
%
% SUPP TABLES
%
% S1 Table
\begin{table}[]
\caption{\textbf{Full organization names}}
\label{table:supp:orglabels}
\end{table}

% S2 Table
\begin{table}[]
\caption{\textbf{L2 Norm of country's representative vectors}.}
\label{table:supp:norm}
\end{table}

@@

%
% SUPP FIGURES
%

%
% Figure - publications over time
%
\begin{figure}[]
	%\includegraphics[width=\textwidth]{\figdir/Descriptive/pubs_over_time.pdf}
	\caption{
		\textbf{Publications over time.}
		}
	\label{fig:supp:pubs_over_time}
\end{figure}
